---
title: "Locations and dive data exploration"
author: "Emily Nazario"
date: "`r Sys.Date()`"
format:
 html: 
  self-contained: true
editor: visual
toc: TRUE
toc-title: "On this page"
theme: yeti
fontcolor: "#134f5c"
execute: 
  message: false
  warning: false
---

```{r}
#| label: Load libraries and read data
#| include: false
library(tidyverse)
library(here)

bor_nestlings <- read_csv(here("data/73_species.csv"))
bor_traits <- read_csv(here("data/Traits_73_species.csv"))
avonet <- readxl::read_excel(here("data/AVONET Supplementary dataset 1.xlsx"), sheet = 2)

```

## Explore 

Seeing if nestlings are showing up earlier in the year over time. First, plot a linear model to see how mean nestling arrival (presented as mean day of year) changes over time. 

```{r}
all_birds_trend <- bor_nestlings %>% 
  group_by(Year) %>% 
  summarize(mean_doy = mean(Dayofyear))

ggplot(all_birds_trend, aes(Year, mean_doy)) + 
  geom_point() +
  geom_smooth(method = "lm")

```

Next, we break it down by species 
```{r}
species_trends <- bor_nestlings %>% 
  group_by(Year, Species) %>% 
  summarize(mean_doy = mean(Dayofyear),
            .groups = "drop")

ggplot(species_trends, aes(Year, mean_doy, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")

```

Amazing! It's a mess! Now we just keep the 5 most data-rich species 
```{r}
data_richness <- bor_nestlings %>% 
  count(Species)

most_rich <- data_richness %>% 
  arrange(desc(n)) %>% 
  slice(1:5)

most_rich_trends <- bor_nestlings %>% 
  filter(Species %in% most_rich$Species) %>% 
  group_by(Species, Year) %>% 
  summarize(mean_doy = mean(Dayofyear), 
            .groups = "drop")

ggplot(most_rich_trends, aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Better -- but one species is not like the others. Which species have the strongest trends?
```{r}
#let's put the FUN in function
# Finds the slope of the relationship between y and x
trend <- function(x, y) {
  xy_lm <- lm(y ~ x)
  coef(xy_lm)[2]
}

# Calculate the trend for all species
bor_trends <- species_trends %>% 
  group_by(Species) %>% 
  summarize(doy_trend = trend(Year, mean_doy))

#Check two species from above DF
soi <- c("ARDCIN", "LARMIN")

species_trends %>% 
  filter(Species %in% soi) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

# My tern (#birds)!

Find the species with the most positive and negative trends, then just plot those 
```{r}
nrow_bor_trends <- nrow(bor_trends) # Use this later
bor_extreme <- bor_trends %>% 
  # Sort by the day of year trend
  arrange(doy_trend) %>%
  # Keep just the first (most negative trend) and last (most positive trend) rows
  slice(c(1, 73))

# Now plot them
species_trends %>% 
  filter(Species %in% bor_extreme$Species) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) + 
    geom_point() +
    geom_smooth(method = "lm")
```

# More data!

Test the hypothesis that smaller birds have more flexible phenology. We need to join our nestling arrival, phenology, and functional traits datasets. 
```{r}
# First, select and rename important columns
avonet_size <- select(avonet, SciName = Species1, Mass_g = Mass)
bor_sciname <- select(bor_traits, 
                      Species = Abbreviation, 
                      SciName = `Scientific name`)

#Join!
bor_trends_size <- bor_trends %>% 
  left_join(bor_sciname, by = "Species") %>%
  left_join(avonet_size, by = "SciName") %>% 
  mutate(abs_trend = abs(doy_trend))

# Plot it
ggplot(bor_trends_size, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")

```

Doesn't look like any strong patterns so far but it's hard to read. Let's modify this to make it easier! 
```{r}
bor_trends_size2 <- bor_trends_size %>%
  drop_na(Mass_g) %>%
  filter(Mass_g < 2000, abs_trend < 1)
  
ggplot(bor_trends_size2, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")
```

**Is our hypothesis supported??** No, the linear model plotted in the figure above does not suggest that smaller birds have more flexible phenologies. 
